Kindly Bent to Free Us

### 2019-07-06 kindly/lang.tex
Write the rest of the explanation for SD typing.

In the rest of this section, we present the syntax (\cref{syntax}),
a linearity-aware semantics (\cref{sem}), syntax directed typing rules
(\cref{sdtyping}) and automatic rules for region annotations (\cref{regionannot}).

@@

\TODO{Do we explain environments here, or push that for typing, later on ?}

@@

We will now present the novel aspects of our system which are necessary to handle
three specific features: binders, borrows, and sharing.

@@

One main feature of our system is to control the usage of variables, either
through
linearity and affinity, which controls how variables are shared or ignored,
or through borrows, which allow to circumvent the linearity rules
in a controlled way.
In order to check these properties, we introduce several new bindings
in our typing environments.
The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
On top of the traditional type and value bindings, noted respectively
$\bvar{\tvar}{\kschm}$ and $\bvar{x}{\schm}$, we also consider two new
bindings.
Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ which indicates
that the borrow $\borrow{x}$ can be used directly, as implemented
by the rule {\sc Borrow} in \cref{selectrules:borrow}.
Suspended bindings, noted $\svar{x}{\schm}^n$, indicates that $x$
cannot be used immediately, but will be usable in a region
marked with $x$ (as detailed in \cref{sdtyp:regions}).

In \lang, many properties are encoded through the use of environment-wide
constraints. Here we present how to control both unused variables and
closure captures through such environment constraints, noted $\Cleq{\E}{k}$.
The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
$\bvar{x}{\tau}$ in $\E$,
if type $\tau$ has kind $k'$ ($\inferSK{C}{\E}{\tau}{k'}$)
then $\Cleq{k'}{k}$.
While this definition is incomplete, it is sufficient to understand
its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
%
The {\sc Var} rule looks up the type scheme of the variable $x$ in
the environment $\E$
and instantiate a concrete type through the substitution $\unif$. It then
checks that the other bindings in $\E$ can safely be discarded by
ensuring they are not linear. This is achieved through the constraint
$\Cleq{\E\Sdel{X}}{\kaff_\infty}$.
%
The {\sc Lam} constraints must ensures that if the body of the lambda captures
linear or affine variable, the content will be reflected in the kind
annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
If one of the binding of $\E$ constraints a type whose kind is affine,
we will derive the constraints $\Cleq{\kaff_n}{k}$, which ensures that
the kind on the arrow is at least affine.

\begin{figure}[!h]
  \begin{minipage}{0.33\linewidth}
    \begin{align*}
      \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
      B &::=\ \bnone \tag{Empty} \\
         &\mid \bvar{x}{\schm}\tag{Variables}\\
         &\mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}\tag{Borrows}\\
         &\mid \svar{x}{\schm}^n \tag{Suspended}\\
         &\mid \bvar{\tvar}{\kschm}\tag{Types}
         % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
    \end{align*}
    \caption{Type environments}
    \label{grammar:env}
  \end{minipage}\hfill
  \begin{minipage}{0.65\linewidth}
    \begin{mathpar}
      \ruleSDBorrow
      \and
      \ruleSDLam
      \and
      \ruleSDVar
    \end{mathpar}
    \caption{The {\sc Borrow} rule}
    \label{selectrules:borrow}
  \end{minipage}
\end{figure}

@@

Linear and affine types also control the copy of variables.
We now showcase how this is achieved in \lang through the {\sc Pair} typing rule
in \cref{sdtyp:pair}.
\TODO{Explain splitting}

@@

    \end{tabular}
    \caption{Splitting rules for bindings}
  \end{subfigure}
\end{figure}

@@

In order to provide good support for imperative programming, \affe allows
to borrow resources.
Such borrows are constrained to a given syntactic region. We note
$\region{x}{e}$ for the region allowing borrows of $x$ to be used in $e$
and where $n$ is the nesting of the region.
Borrows can not be used outside of a corresponding region.
These remarks lead to an
algorithm for automatic region annotation: start from each borrow, create
a region, and extend it as much as possible while respecting other borrows
and binders.
A complete specification of the algorithm is provided in \cref{regionannot}.
In the rest of this section, we will only consider terms
with fully annotated regions.

\TODO{Transition}

In order to use a suspended binding for the variable $x$,
we need to enter a \emph{region} associated with $x$,
noted $\region{x}{e}$.
The main idea of region typing (implemented by the {\sc Region} rule)
is to replace suspended bindings by borrow bindings, typecheck the body
of the region, and finally ensure that the borrow made available inside
the region does not leak outside.
This last check is done by using integer indices that correspond to the nesting
level of the region. The kind of borrow is annotated with the integer
corresponding to their region, and a final constraint ensure that
the return type of the region does not contain such borrow.

We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{env:rule:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ for our example and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).

@@

\subsubsection{Constraints}

>>>>

In the rest of this section, we present the syntax (\cref{syntax})
the syntax directed typing rules (\cref{sdtyping})
and a linearity-aware semantics (\cref{sem}) for \lang.

@@

@@

\subsubsection{Kinds and constraints}

One novel feature or \affe is the use of kinds and qualified types to encode
linear and affine types. Before explaining the details of our type system, we
give a quick presentation of the kinds and constraints.
A kind $k$ can be either a kind variable, $\kvar$, or a constant.
Each constant is composed of a ``quality'' and a ``level''.
The quality describes the use pattern of the type and can be
either unrestricted $\kun$, affine $\kaff$ or linear $\klin$. The level
describes the nested regions in which it can be used. Levels are represented
by a natural number of $\infty$.
$0$ represents the normal scope outside of any region and is often omitted
(for instance $\kaff_0$ is simply noted $\kaff$).
$\infty$ is the empty region that is infinitely nested and is mostly used
as an upper bound.
For instance, the constraint $\Cleq{\kvar}{\kun_\infty}$ indicates that
$\kvar$ must be unrestricted, but can be local to a region.
%
Just as presented before, constraints present in schemes and type signatures
are simply a list of atomic constraints, which are composed
of kind inequalities.
%
Our inequality constraints are based on an inequality relation on
kind constants noted $\lk_\Lat$. This relation forms a lattice
and  is summarized in
\cref{sdtyp:lattice}. We have the basic inequalities on qualities:
$\kun_n \lk_\Lat \kaff_n \lk_\Lat \klin_n$ for arbitrary regions $n$.
Additionally, the inequality follows the region nesting, for
instance $\kaff_n \lk_\Lat \kaff_{n+1}$.

\begin{figure*}[!h]
  \input{infer/lattice}
  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
  \label{sdtyp:lattice}
\end{figure*}

@@

\lang aims to control the usage of variables, either
through
linearity and affinity which restrict how variables are shared and ignored,
or through borrows which allow to circumvent the linearity rules
in a controlled way.
In order to check these properties, we introduce several new bindings
in our typing environments.
The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
On top of the traditional type and value bindings, noted respectively
$\bvar{\tvar}{\kschm}$ and $\bvar{x}{\schm}$, we also consider two new
bindings.
Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ indicates
that the borrow $\borrow{x}$ can be used directly, as implemented
by the rule {\sc Borrow} in \cref{selectrules:borrow}.
Suspended bindings, noted $\svar{x}{\schm}^n$, indicates that $x$
cannot be used immediately, but will be usable in a region
marked with $x$, which we detail in \cref{sdtyp:regions}.

Environments are also used to encode several linearity-related properties
through \emph{environment-wide constraints}.
Here we present how to control both unused variables and
closure captures through such environment constraints, noted $\Cleq{\E}{k}$.
The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
$\bvar{x}{\tau}$ in $\E$,
if type $\tau$ has kind $k'$ (noted $\inferSK{C}{\E}{\tau}{k'}$)
then the following constraints should hold: $\Cleq{k'}{k}$.
While this definition is incomplete, it is sufficient to understand
its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
%
The {\sc Var} rule looks up the type scheme of the variable $x$ in
the environment $\E$
and instantiate a concrete type through the substitution $\unif$. It then
checks that the other bindings in $\E$ can safely be discarded by
ensuring they are not linear. This is achieved through the constraint
$\Cleq{\E\Sdel{x}}{\kaff_\infty}$ which replaces the traditional
``weakening'' rule found in many type systems.

The {\sc Abs} rule must ensures that if the body of the lambda captures
linear or affine variables, the content will be reflected in the kind
annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
If one of the binding of $\E$ contains a type whose kind is affine,
we will derive the constraints $\Cleq{\kaff}{k}$, which ensures that
the kind on the arrow is at least affine.
Note that $x$ is \emph{not} included in $\E$: since it's the argument of the
closure, its use should not induce a capture.

\begin{figure}[!h]
  \begin{minipage}{0.33\linewidth}
    \begin{align*}
      \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
      B &::=\ \bnone \tag{Empty} \\
         &\mid \bvar{x}{\schm}\tag{Variables}\\
         &\mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}\tag{Borrows}\\
         &\mid \svar{x}{\schm}^n \tag{Suspended}\\
         &\mid \bvar{\tvar}{\kschm}\tag{Types}
         % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
    \end{align*}
    \caption{Type environments}
    \label{grammar:env}
  \end{minipage}\hfill
  \begin{minipage}{0.65\linewidth}
    \begin{mathpar}
      \ruleSDBorrow
      \and
      \ruleSDLam
      \and
      \ruleSDVar
    \end{mathpar}
    \caption{The {\sc Borrow}, {\sc Abs} and {\sc Var} rules}
    \label{selectrules:borrow}
    \label{selectrules:binders}
  \end{minipage}
\end{figure}

@@

Linear and affine types also control the copy of variables.
To demonstrate how this is achieve in \lang, we show the
{\sc Pair} typing rule in \cref{sdtyp:pair}.
In the {\sc Pair} rule, we split the type environment $\E$ in two parts,
$\E_1$ and $\E_2$, which are used to typecheck the components of the pair.
We note the splitting $\lsplit{C}{\E}{\E_1}{\E_2}$ where $C$ is a constraint
that must be respected for the split to be valid.
The main splitting rules are shown in \cref{sdtyp:split}. Each
rule applies to a binding for a given variable. For instance, the
{\sc Both} rule indicates that if we can prove that $x$ is unrestricted, we
can duplicate it in both environments.
Similarly, since immutable borrows are always unrestricted, the
{\sc Borrow} rule allows to duplicate them arbitrarily.
The {\sc Left} and {\sc Right} rule allow to move a binding
to the left or right environment.
The {\sc Susp} and {\sc SuspB} allow to split off suspended bindings to
the left environment while conserving access to the binding on the right.
Such suspended binding can later be turned
into a borrow inside a region. The asymmetry here must always follow
the order of execution of the language: in \affe, execution is from left to
right, which means that a resource can be used as a borrow on the left
and then as a full resource on the right, assuming that the borrow does not
escape. The {\sc Susp} rule work with a full resource while the {\sc SuspB}
rule allow to split off an immutable suspended binding based off a borrow.

Using these various rules, we can distribute the bindings every time an
expression has multiple sub-expressions such as pairs, function applications
and let bindings. For instance, in the
expression
$\letin{a}{\text{create}\ 10\ x}
{\introPair{\region[{}]{x}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
we use the rule {\sc Susp} to detach a borrow on the left of the pair.
Naturally, we must here ensure that the borrow on the left of the pair does not
escape, which brings us to the typing rules for regions.

@@

    \end{tabular}
    \caption{Splitting rules for bindings}
    \label{sdtyp:split}
  \end{subfigure}
\end{figure}

@@

In order to provide good support for imperative programming, \affe allows
to borrow resources.
Such borrows are constrained to a given syntactic region. We note
$\region{x}{e}$ the region which allow borrows of $x$ to be used in $e$
and where $n$ is the nesting of the region.
Borrows can not be used outside of a corresponding region.
These remarks lead to an
algorithm for automatic region annotation: start from each borrow, create
a region, and extend it as much as possible while respecting other borrows
and binders.
A complete specification of the algorithm is provided in \cref{regionannot}.
In the rest of this section, we will only consider terms
with fully annotated regions.

The main idea of region typing (implemented by the {\sc Region} rule)
is to replace suspended bindings by borrow bindings, typecheck the body
of the region, and finally ensure that the borrow made available inside
the region does not leak outside.
This last check is done by using integer indices that correspond to the nesting
level of the region. The kind of borrow is annotated with the integer
corresponding to their region, and a final constraint ensure that
the return type of the region does not contain such borrow.

We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{selectrules:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ for our example and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).

@@
